---
title: "Econ 104 Project 3"
author: "Group 12"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      out.width = "75%", fig.align = "center")
```

\pagebreak

```{r}
# Libraries
library(AER)
library(plm)
library(gplots)
library(tidyr)
library(broom)
library(knitr)
library(margins)
```


# 1. Panel Data Model
```{r}
# Loading a panel data
data("OECDGas")
```

## (a) Briefly discuss your data and the question you are trying to answer with your model.
### Ciation
```{r}
# Data citation
citation(package = "AER")
```

### Description of the data
```{r}
paste("N =", length(unique(OECDGas$country))) # number of individuals
paste("T =", length(unique(OECDGas$year))) # number of years
```

```{r}
# subsetting 5 countries for the project purpose
oecd_gas <- OECDGas[OECDGas$country == "Spain"|
                      OECDGas$country == "Italy"| 
                      OECDGas$country == "Japan"|
                      OECDGas$country == "Netherlands"|
                      OECDGas$country == "Greece",]

oecd_gas$year <- factor(oecd_gas$year)
oecd_gas$country <- factor(as.character(oecd_gas$country))

```

The dataset we chose, `OECDGas`, is a ***balanced panel data*** on gasoline consumption in 18 OECD countries over 19 years, spanning from 1960-1978.

As the original panel data has 18 individuals (N; countries) observed for 19 years (T), the original panel data is considered *wide and long*. However, for this project purpose, we selected 5 countries with relatively high average gasoline consumption over the years, which include `Spain`, `Italy`, `Japan`, `Netherlands`, `And Greece`. After reducing the number of individuals, we are now dealing with **Long and Narrow** panel data, where N = 5 and T = 19.

In this first part of the project, the question we attempt to answer is: 

***"Within the same time frame, how do the impacts of per-capita income, gasoline prices, and the number of cars on motor gasoline consumption differ across countries?"***

In other words, we are attempting to identify whether or not the impacts of the independent variables are statistically different across countries, and if so, which model is the most appropriate to capture the dynamics, among pooled, fixed effects, and random effects models.

For more detailed descriptions of the variables, refer to part (b).


## (b) Provide a descriptive analysis of your variables. This should include relevant figures with comments including some graphical depiction of individual heterogeneity.


### Dependent Variable

#### `gas`: 
Logarithm of motor gasoline consumption per car. 
Our dependent variable, `gas`, appears to be right-skewed.
```{r}
par(mfrow = c(1,2))
hist(oecd_gas$gas,
     main = "Distribution of Gas", prob = TRUE,
     xlab = "Log-Motor gasoline consumption",
     col = "purple")
lines(density(oecd_gas$gas), col = "purple4", lwd = 2)
boxplot(oecd_gas$gas, horizontal = TRUE,
        main = "Boxplot of Gas", col = "plum")

```


  
### Independent Variable

#### `income`: 
Logarithm of real per-capita income. 
`income` variable is slightly left-skewed.
```{r}
par(mfrow = c(1,2))
hist(oecd_gas$income,
     main = "Distribution of Income", prob = TRUE,
     xlab = "Log-real per-capita income",
     col = "red")
lines(density(oecd_gas$income), col = "red4", lwd = 2)
boxplot(oecd_gas$income, horizontal = TRUE,
        main = "Boxplot of Income", col = "salmon")

```



#### `price`:
Logarithm of real motor gasoline price. 
`price` is right-skewed.
```{r}
par(mfrow = c(1,2))
hist(oecd_gas$price,
     main = "Distribution of Prices", prob = TRUE,
     xlab = "Log-real motor gasoline price",
     col = "blue")
lines(density(oecd_gas$price), col = "blue4", lwd = 2)
boxplot(oecd_gas$price, horizontal = TRUE,
        main = "Boxplot of Prices", col = "lightblue")

```



#### `cars` : 
Logarithm of the stock of cars per-capita. 
`cars` is left-skewed.
```{r}
par(mfrow = c(1,2))
hist(oecd_gas$cars,
     main = "Distribution of Cars", prob = TRUE,
     xlab = "Log-stock of cars per-capita.",
     col = "green")
lines(density(oecd_gas$cars), col = "green4", lwd = 2)
boxplot(oecd_gas$cars, horizontal = TRUE,
        main = "Boxplot of Cars", col = "lightgreen")

```







```{r}
# Converting the data into a panel data
pd_gas <- pdata.frame(oecd_gas, index = c("country", "year"))
```


## (c) Fit the three models below, and identify which model is your preferred one and why. Make sure to include your statistical diagnostics to support your conclusion, and to comment on your findings.

#### Exploring Heterogeneity across Countries

```{r}
# Heterogeneity across countries
boxplot(oecd_gas$gas ~ oecd_gas$country, 
        col = c("red","orange","yellow","green", "blue" ),
        xlab = "Country",
        ylab = "Log-Gas consumption")
plotmeans(gas ~ country, data = oecd_gas)

```

As shown above, heterogeneity across individuals exists. `Greece`, `Italy`, `Netherlands`, and `Spain` seem to be similarly spread out, however, `Japan` seems to have a higher variance compared to the others. While we can't conclude anything statistically at this point, it still provides a good intuitive sense that each country may not be constant.

#### Exploring Heterogeneity across Time

```{r}
# Creating a time variable
t <- as.numeric(as.character(oecd_gas$year))

# Creating a color vector
color_gradient <- colorRampPalette(c("lightblue", "darkblue"))(length(unique(t)))
color_vector <- rep(color_gradient, 4)

# Note outliers were remove for clear visualization
boxplot(gas ~ year, data = oecd_gas, outline = FALSE, col = color_vector)
plotmeans(gas ~ year, data = oecd_gas, col = color_vector)


```

It is observed that heterogeneity also seems to exist across time. The variability tends to decrease for the more recent years, and there is a gradual downward trend.


#### Visualizing Two-way Heterogeneity:
\
The three plots below visualize the relationship between the dependent variable and each independent variable.
As shown below, `gas` and `income`, and `gas` and `cars` seem to be negatively correlated, and the relationship between `gas` and `price` is not clear based purely on the plot. 

The scatter plots also include color effects where darker colors are more recent years. By visually inspecting the two-way heterogeneity, we can get a big picture that there may exist both individual and time heterogeneity.
```{r}
# Creating a time variable
t <- as.numeric(as.character(oecd_gas$year))

# Creating a color vector
color_gradient <- colorRampPalette(c("lightblue", "darkblue"))(length(unique(t)))
color_vector <- rep(color_gradient, 4)


color_gradient1 <- colorRampPalette(c("lightpink", "darkred"))(length(unique(t)))
color_vector1 <- rep(color_gradient1, 4)


color_gradient2 <- colorRampPalette(c("lightgreen", "darkgreen"))(length(unique(t)))
color_vector2 <- rep(color_gradient2, 4)


# Gas vs. income
plot(oecd_gas$income, oecd_gas$gas, col = color_vector, pch = 20,
     main = "Motor Gasoline Consumption vs. Per-Capita Income",
     xlab = "Log-Per-Capita Income", ylab = "Log-Motor Gasoline Consumption")
legend("topright", legend = unique(oecd_gas$year), 
       col = color_gradient, pch = 20, cex = 0.5)

# Gas vs. price
plot(oecd_gas$price, oecd_gas$gas, col = color_vector1, pch = 20,
     main = "Motor Gasoline Consumption vs. Price",
     xlab = "Log-Per-Capita Income", ylab = "Log-Pirce of Gasoline")
legend("topright", legend = unique(oecd_gas$year), 
       col = color_gradient1, pch = 20, cex = 0.5)

# Gas vs. cars
plot(oecd_gas$cars, oecd_gas$gas, col = color_vector2, pch = 20,
     main = "Motor Gasoline Consumption vs. Number of Cars",
     xlab = "Log-Per-Capita Income", ylab = "Log-Stock of Cars Per-Capita")
legend("topright", legend = unique(oecd_gas$year), 
       col = color_gradient2, pch = 20, cex = 0.5)


```

#### Correlation Matrix:
\
Based on the correlation matrix, we can see the dependent variable and the independent variables are negatively correlated, which was also expected as we explored the visuals above. Particularly, the `income` and `cars` variables are strongly correlated with the dependent variable.
```{r}
library(corrplot)
corrplot.mixed(cor(oecd_gas[3:6]),lower = "pie", upper = "color", addCoef.col="black")

```

#### The Big Picture

The plot below visualizes the motor gasoline consumption data for the five countries. We can see that `Japan` has a little steeper slope than other countries, but overall, a slow and steady downward trend is observed, which verifies again that there exists heterogeneity across time.

```{r}
# Converting the year variable to numeric
oecd_gas$year <- as.numeric(as.character(oecd_gas$year))



plot(gas ~ year, data = oecd_gas, pch = 20, main = "Moter Gasoline Consumption per car")
lines(unique(oecd_gas$year), 
      oecd_gas$gas[oecd_gas$country == unique(oecd_gas$country)[1]], col = "red")
lines(unique(oecd_gas$year), 
      oecd_gas$gas[oecd_gas$country == unique(oecd_gas$country)[2]], col = "orange")
lines(unique(oecd_gas$year), 
      oecd_gas$gas[oecd_gas$country == unique(oecd_gas$country)[3]], col = "yellow")
lines(unique(oecd_gas$year), 
      oecd_gas$gas[oecd_gas$country == unique(oecd_gas$country)[4]], col = "green")
lines(unique(oecd_gas$year), 
      oecd_gas$gas[oecd_gas$country == unique(oecd_gas$country)[5]], col = "blue")
legend("topright", 
       legend = unique(oecd_gas$country), 
       lty = 1, col = c("red","orange","yellow","green", "blue"))

```

\pagebreak

### • ***Pooled Model***

The pooled model is a model that assumes constant marginal effects across individuals. In other words, the model has no provision for individual differences that could potentially result in different coefficients. The mathematical notation is as follows:

$$Gas_{it}  = \beta_1 + \beta_2Income_{it} + \beta_3 Price_{it} + \beta_4 Cars_{it} + \epsilon_{it}$$

Note that the coefficients do NOT have *i* or *t* subscripts, which implies:

  1. $\beta$ does not change over time
  2. $\beta$ does not change across individuals

Below are the summary statistics and plots that provide detailed information on our pooled model:

```{r}
# Pooled Model --> No indiv. diff. (same coef. across indiv.)
# Estimating the pooled OLS model
gas_pool <- plm(gas ~ income + price + cars, model = "pooling", 
                data = pd_gas)

summary(gas_pool)

```

```{r} 
# Trying to output the fitted values manually
a <- gas_pool$coefficients[[1]]
b1 <- gas_pool$coefficients[[2]]
b2 <- gas_pool$coefficients[[3]]
b3 <- gas_pool$coefficients[[4]]


Y <- oecd_gas$gas
x1 <- oecd_gas$income
x2 <- oecd_gas$price
x3 <- oecd_gas$cars

y_fitted <- a + b1*x1 + b2*x2 + b3*x3


df_fitted <- data.frame(country = oecd_gas$country, y_fitted = y_fitted)

```

#### Visualizing the model fit for each country
```{r}
# Fitted values for each country (pooled model)
greece_fitted <- df_fitted[df_fitted$country == 
                             unique(oecd_gas$country)[1],]$y_fitted
italy_fitted <- df_fitted[df_fitted$country == 
                            unique(oecd_gas$country)[2],]$y_fitted
japan_fitted <- df_fitted[df_fitted$country == 
                            unique(oecd_gas$country)[3],]$y_fitted
netherlands_fitted <- df_fitted[df_fitted$country == 
                                  unique(oecd_gas$country)[4],]$y_fitted
spain_fitted <- df_fitted[df_fitted$country == 
                            unique(oecd_gas$country)[5],]$y_fitted
```

```{r}
unique_country <- unique(oecd_gas$country)


par(mfrow = c(2,3))
# Greece
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[1]],
     main = "Greece", ylim = c(4, 6), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), greece_fitted, col = "red")

#Italy
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[2]],
     main = "Italy", ylim = c(2.5, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), italy_fitted, col = "orange")

# Japan
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[3]],
     main = "Japan", ylim = c(3.5, 6.5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), japan_fitted, col = "yellow")

# Netherlands
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[4]],
     main = "Netherlands", ylim = c(3, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), netherlands_fitted, col = "green")

# Spain
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[5]],
     main = "Spain", ylim = c(3, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), spain_fitted, col = "blue")

# Combined
plot(gas ~ year, data = oecd_gas, main = "Combined")
lines(unique(oecd_gas$year), greece_fitted, col = "red")
lines(unique(oecd_gas$year), italy_fitted, col = "orange")
lines(unique(oecd_gas$year), japan_fitted, col = "yellow")
lines(unique(oecd_gas$year), netherlands_fitted, col = "green")
lines(unique(oecd_gas$year), spain_fitted, col = "blue")


```

#### Residuals Plot
```{r}
df_resid <- data.frame(country = oecd_gas$country, 
                       resid =  as.numeric(gas_pool$residuals))

par(mfrow = c(2,3))
plot(greece_fitted, df_resid$resid[df_resid$country == unique_country[1]], 
     main = "Residuals Plot (Greece)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(italy_fitted, df_resid$resid[df_resid$country == unique_country[2]], 
     main = "Residuals Plot (Italy)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(japan_fitted, df_resid$resid[df_resid$country == unique_country[3]], 
     main = "Residuals Plot (Japan)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(netherlands_fitted, df_resid$resid[df_resid$country == unique_country[4]], 
     main = "Residuals Plot (Netherlands)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(spain_fitted, df_resid$resid[df_resid$country == unique_country[5]], 
     main = "Residuals Plot (Spain)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# As a whole
plot(as.numeric(fitted(gas_pool)),gas_pool$residuals,
     main = "Residuals Plot (Combined)",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```

#### BP (Breusch-Pagan) test
```{r}
# H0: Variance of error term is constant
bptest(gas_pool) # reject H0 --> Heteroskedasticity exists
```


As also can be anticipated by the model fit, the residuals plot and BP test confirm that there exists heteroskedasticity in the residuals, which means the standard errors output earlier are incorrect. Thus, we correct the standard errors using Cluster-robust standard errors.


#### Correcting the standard errors with Cluster-robus standard errors.

```{r}
# Correcting the standard errors using Cluster-robust standard errors
coeftest(gas_pool, vcov = vcovHC(gas_pool,
                                 type = "HC0", cluster = "group"))
```

To summarize, we observed that while the fit describes the actual data relatively well for some countries, such as the `Netherlands`, it is also clear that the fit does not apply to all countries as seen in `Greece` or `Italy`. This pooled model will be compared with a fixed effects model in the next section.


\pagebreak
### • ***Fixed Effects Model***

The fixed effects model relaxes the assumption that was considered in the pooled model, which is that all individuals have the same coefficients. This implies that the coefficients ($\beta$) can be different across individuals. The mathematical notation for the fixed effects model is as follows:

$$Gas_{it}  = \beta_{1i} + \beta_2Income_{it} + \beta_3 Price_{it} + \beta_4 Cars_{it} + \epsilon_{it}$$

Note that an *i* subscipt was added to $\beta_1$. It is assumed that ***individual heterogeneity***, all behavioral differences across individuals are captured by the varying intercept.

In this section, two approaches are used to estimate the fixed effects model.

  1. The Least Squares Dummy Variable Estimator (LSDV)
  2. The Within Estimator


#### 1. LSDV model

For LSDV model, a dummy variable is included for each individual:


$$
\begin{aligned}
D_{Greece,i} &= \begin{cases} 1 & \text{if } i = 1 \\ 0 & \text{otherwise} \end{cases}
&D_{Italy,i} = \begin{cases} 1 & \text{if } i = 2 \\ 0 & \text{otherwise} \end{cases}\\
D_{Japan,i} &= \begin{cases} 1 & \text{if } i = 3 \\ 0 & \text{otherwise} \end{cases}
&D_{Netherlands,i} = \begin{cases} 1 & \text{if } i = 4 \\ 0 & \text{otherwise} \end{cases}\\
D_{Spain,i} &= \begin{cases} 1 & \text{if } i = 5 \\ 0 & \text{otherwise} \end{cases}
\end{aligned}
$$

Thus, the full model would be:

$$
\beta_{11}D_{Greece,i} + \beta_{12}D_{Italy,i} + \beta_{13}D_{Japan,i} + \beta_{14}D_{Netherlands,i} + \beta_{15}D_{Spain,i} +\\
\beta_2Income_{it} + \beta_3 Price_{it} + \beta_4 Cars_{it} + \epsilon_{it}
$$


Below are the summary statistics for the Fixed effects model, outputted by the LSDV approach:


#### Estimating the Fixed Effects model
```{r}
# Fixed Effects -- dummy variables
pd_gas5 <- pdata.frame(oecd_gas[oecd_gas$country %in% 
                                  c("Greece", "Italy", "Japan", "Netherlands", "Spain"),])

# LSDV approach
gas_fixed <- lm(gas ~ income + price + cars + factor(country) - 1, data = pd_gas5)

summary(gas_fixed)



```


#### 2. Within Approach
```{r}
# Estimating the within model
gas_within <- plm(gas ~ income + price + cars,
                  model = "within",data= pd_gas)

# Extracting the fixed effect for each country
fixef(gas_within)
```


#### Visual inspection
```{r}
df_fixed <- data.frame(country = oecd_gas$country, fitted = fitted(gas_fixed))
```

```{r}
# Fitted values for each country (Fixed)
greece_fixed <- df_fixed[df_fixed$country == unique(oecd_gas$country)[1],]$fitted
italy_fixed <- df_fixed[df_fixed$country == unique(oecd_gas$country)[2],]$fitted
japan_fixed <- df_fixed[df_fixed$country == unique(oecd_gas$country)[3],]$fitted
netherlands_fixed <- df_fixed[df_fixed$country == unique(oecd_gas$country)[4],]$fitted
spain_fixed <- df_fixed[df_fixed$country == unique(oecd_gas$country)[5],]$fitted
```

```{r}
par(mfrow = c(2,3))
# Greece
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[1]],
     main = "Greece", ylim = c(4, 6), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), greece_fixed, col = "red")

#Italy
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[2]],
     main = "Italy", ylim = c(2.5, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), italy_fixed, col = "orange")

# Japan
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[3]],
     main = "Japan", ylim = c(3.5, 6.5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), japan_fixed, col = "yellow")

# Netherlands
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[4]],
     main = "Netherlands", ylim = c(3, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), netherlands_fixed, col = "green")

# Spain
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[5]],
     main = "Spain", ylim = c(3, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), spain_fixed, col = "blue")

# Combined
plot(gas ~ year, data = oecd_gas, main = "Combined")
lines(unique(oecd_gas$year), greece_fixed, col = "red")
lines(unique(oecd_gas$year), italy_fixed, col = "orange")
lines(unique(oecd_gas$year), japan_fixed, col = "yellow")
lines(unique(oecd_gas$year), netherlands_fixed, col = "green")
lines(unique(oecd_gas$year), spain_fixed, col = "blue")


```

#### Residuals Plot
```{r}
df_resid_fixed <- data.frame(country = oecd_gas$country, 
                             resid =  as.numeric(gas_fixed$residuals))

par(mfrow = c(2,3))
plot(greece_fixed, df_resid_fixed$resid[df_resid_fixed$country == unique_country[1]], 
     main = "Residuals Plot (Greece)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(italy_fixed, df_resid_fixed$resid[df_resid_fixed$country == unique_country[2]], 
     main = "Residuals Plot (Italy)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(japan_fixed, df_resid_fixed$resid[df_resid_fixed$country == unique_country[3]], 
     main = "Residuals Plot (Japan)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(netherlands_fixed, df_resid_fixed$resid[df_resid_fixed$country == unique_country[4]], 
     main = "Residuals Plot (Netherlands)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(spain_fixed, df_resid_fixed$resid[df_resid_fixed$country == unique_country[5]], 
     main = "Residuals Plot (Spain)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# As a whole
plot(as.numeric(fitted(gas_fixed)),gas_fixed$residuals,
     main = "Residuals Plot (Combined)",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

```


#### Hypothesis testing:

As seen above, we can see improvement in the fit using the fixed effects model. Now, we verify if the improvement observed is statistically significant. In other words, using diagnostic statistics, we can confirm if the fixed effects are actually needed to model the original panel data.

```{r}
# Hypothesis test; H0: no fixed effects needed
pFtest(gas_within, gas_pool)
```

The p-value is very close to zero, which means we reject the null hypothesis that the heterogeneity across individuals is not statistically significant. **Therefore, we can conclude that the fixed effects are necessary to account for individual-specific effects or time-variant characteristics that significantly contribute to the variation in the panel data.**

\pagebreak
### • ***Random Effects Model***

In the random effects model, it is assumed that the individual differences are **random** as opposed to the fixed effects model, where we consider them **fixed**. One similarity is that the random effects are also captured in the intercept parameter ($\beta_{1i}$), which consists of two parts: fixed and random. 

$$\beta_{1i} = \bar\beta_1 + u_i$$
where 

**$\bar\beta_1$**: population mean (fixed)

**$u_i$**: random effects (random)

$$Gas_{it}  = \bar\beta_1 + \beta_2Income_{it} + \beta_3 Price_{it} + \beta_4 Cars_{it} + v_{it}$$
and the error term $v_{it}$ is:

$$v_{it} = u_i + e_{it}$$

#### Estimating the random effect model
```{r}
# Estimating the random effect model
gas_random <- plm(gas ~ income + price + cars,
                  model = "random", data = pd_gas)
summary(gas_random)
```

#### Visual Inspection
```{r}
df_random <- data.frame(country = oecd_gas$country, fitted = fitted(gas_random))
```

```{r}
# Fitted values for each country (Random)
greece_random <- df_random[df_random$country == unique(oecd_gas$country)[1],]$fitted
italy_random <- df_random[df_random$country == unique(oecd_gas$country)[2],]$fitted
japan_random <- df_random[df_random$country == unique(oecd_gas$country)[3],]$fitted
netherlands_random <- df_random[df_random$country == unique(oecd_gas$country)[4],]$fitted
spain_random <- df_random[df_random$country == unique(oecd_gas$country)[5],]$fitted
```

```{r}
par(mfrow = c(2,3))
# Greece
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[1]],
     main = "Greece", ylim = c(min(y_fitted), 6), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), greece_random, col = "red")

#Italy
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[2]],
     main = "Italy", ylim = c(2.5, 4.5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), italy_random, col = "orange")

# Japan
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[3]],
     main = "Japan", ylim = c(min(y_fitted), 6), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), japan_random, col = "yellow")

# Netherlands
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[4]],
     main = "Netherlands", ylim = c(2.5, 5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), netherlands_random, col = "green")

# Spain
plot(unique(oecd_gas$year), oecd_gas$gas[oecd_gas$country == unique_country[5]],
     main = "Spain", ylim = c(2.5, 5.5), xlab = "Date", ylab = "Gas")
lines(unique(oecd_gas$year), spain_random, col = "blue")

# Combined
plot(gas ~ year, data = oecd_gas, main = "Combined", ylim = c(2.5, 6.5))
lines(unique(oecd_gas$year), greece_random, col = "red")
lines(unique(oecd_gas$year), italy_random, col = "orange")
lines(unique(oecd_gas$year), japan_random, col = "yellow")
lines(unique(oecd_gas$year), netherlands_random, col = "green")
lines(unique(oecd_gas$year), spain_random, col = "blue")


```

From the visual inspection, we can see that the model tends to underestimate the true values. Compared to the fixed effects model, the model fit of the random effects model indicates inaccuracy and slight biases.

#### Residuals Plot
```{r}
df_resid_random <- data.frame(country = oecd_gas$country, 
                              resid =  as.numeric(gas_random$residuals))

par(mfrow = c(2,3))
plot(greece_random, df_resid_random$resid[df_resid_random$country 
                                          == unique_country[1]], 
     main = "Residuals Plot (Greece)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(italy_random, df_resid_random$resid[df_resid_random$country 
                                         == unique_country[2]], 
     main = "Residuals Plot (Italy)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(japan_random, df_resid_random$resid[df_resid_random$country 
                                         == unique_country[3]], 
     main = "Residuals Plot (Japan)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(netherlands_random, df_resid_random$resid[df_resid_random$country 
                                               == unique_country[4]], 
     main = "Residuals Plot (Netherlands)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
plot(spain_random, df_resid_random$resid[df_resid_random$country 
                                         == unique_country[5]], 
     main = "Residuals Plot (Spain)",
       xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# As a whole
plot(as.numeric(fitted(gas_random)),gas_random$residuals,
     main = "Residuals Plot (Combined)",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)
```


Since we have confirmed earlier that heterogeneity exists across individuals, we decide which model is better by comparing the fixed effects model and the random effects model.


#### Comparing the `Fixed effects model` and `Random effects model`

To decide which model is better, the **Hausman Test** was used, which detects the correlation between the error terms and the regressors (right-hand side of the equation). 
```{r}
# H0: Random Effect model is preferred
# H1: Fixed Effect model is preferred
phtest(gas_within, gas_random)
```

Since the p-value is small, we reject the null hypothesis that the random effects model is preferred to the fixed effects model.


### (d) Conclusion
In conclusion, in the first part of this project, we investigated panel data on gasoline consumption in 5 countries: Greece, Italy, Japan, Netherlands, and Spain. We have explored heterogeneity across individuals (countries) and time and concluded that each country's gas consumption is impacted differently by each variable, including `income`, `price`, and `cars`, and the **fixed effects** across individuals are statistically significant. This answers the question that we aimed to answer; ***"Within the same time frame, how do the impacts of per-capita income, gasoline prices, and the number of cars on motor gasoline consumption differ across countries?"***

